Likelihoods for a general class of ARGs under the SMC

### 2025-02-21 paper.tex
Switch to abbrvnat
Intro para on uncertainty

\section*{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now strong interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024,nielsen_inference_2025}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Numerous applications taking advantage of this detailed history
in inferred ARGs are now appearing~\citep{
stern_approximate_2019,
osmond_estimating_2021,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
ignatieva_ongoing_2022,
wang_complex_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
huang_estimating_2024,
grundler_geographic_2024,
korfmann_simultaneous_2024,
deraje_inferring_2024,
speidel2025high}
and it seems likely that many more will
follow~\citep{harris_database_2019,harris_using_2023}.
Although there are many different approaches to ARG
inference~\citep{wong_general_2023}, two broad classes of method
have been the focus of most recent interest.

% TODO Maybe swap the order of these two paragraphs around, so it's 
% more natrual to talk about the advantages of model-driven stuff next.
The first class of inference methods are Bayesian approaches
that sample ARGs under a population genetics model such as the
coalescent with recombination~\citep{hudson_properties_1983},
and its approximatation, the Sequentially Markovian
Coalescent~\citep{mcvean_approximating_2005,marjoram_fast_2006}.
ARGweaver~\citep{rasmussen_genome-wide_2014} has been the most widely used and studied,
and until recently consistently out-performed other methods in terms 
of accuracy in a variety of benchmarks~\citep{brandt2022evaluation},
and has succesfully been applied in several
contexts~\citep[e.g.,][]{de_chimpanzee_2016,shriner_whole_2018,hejase_genomic_2020,
stankowski_genetic_2024}. 
The recently-introduced method SINGER takes a similar approach,
and by using an improved Monte Carlo algorithm
promises to be even more accurate while being substantially more
scalable~\citep{deng_robust_2024}.

The second class of inference method that has been of recent interest 
is based on heuristics and approximate models.
Relate~\citep{speidel_method_2019},
tsinfer~\citep{kelleher_inferring_2019},
ARG-Needle~\citep{zhang_biobank-scale_2023} 
and Threads~\citep{gunnarsson_scalable_2024}
work on quite different 
principles, but share some common properties. 
Firstly, they are all 
heuristic and approximation driven, favouring computational
scalability over a basis in a statistical model.
Secondly, they all produce a single
ARG as output, inferring a deterministic point estimate with no means 
of quantifying uncertainty. Thirdly (and most importantly for the 
purposes of this paper) the ARGs they produce 
lack some of the information present in ARGweaver or SINGER. 
This occurs, roughly speaking, because of a fundamental difference in approach:
not all details of the ARG can be reliably inferred
(e.g., the timing of recombination events, as discussed below);
and the heuristic methods deal with this uncertainty by omitting these events,
while ARGweaver and SINGER provide a sample from the posterior.
(So, some aspects of the resulting ARGs should be interpreted slightly differently;
see \citet{wong_general_2023} for more discussion.)

@@

Computing the likelihood of an ARG under a model is a fundamental element of
probabilistic inference. The classical Kuhner-Yamato-Felsenstein
\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
works by assigning a probablility to each recombination and common
ancestor event, and to the inter-event  waiting times. The likelihood of an 
ARG is then a product over all these three classes of term.
% In essence, computing the KYF likelihood requires
% ``replaying'' the  state of Hudson's simulation
% algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
% compute the exact configuration of lineages and distribution
% of ancestral material at each time point~\citep{baumdicker_efficient_2021}.
% It therefore requires complete
% detail about each of these events to be stored. 
% Other approaches to defining the likelihood of an ARG under a coalescent-like 
% model follow the same basic event-by-event 
% strategy~\citep[e.g.,][]{guo_recombination-aware_2022}.
% % [Are there others?  The bacterial stuff from Didelot etc seem to focus on
% % a tree-by-tree likelihood, rather than global graph, so not clear it's the 
% % same thing]
% Computing the likelihood under the SMC rather than the Hudson coalescent 
% simplifies bookkeeping somewhat,  % Dunno what I'm talking about here
% but a complete event-by-event description of the 
% input ARG is still required.
Although the KYF formulation is a natural and elegant way to describe 
the likelihood of an ARG, such complete detail is not present 
in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
Simulators such as the classical \texttt{ms}
program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
which does not contain sufficient detail for likelihood computation.
The \texttt{msprime}
simulator~\citep{kelleher_efficient_2016,baumdicker_efficient_2021} 
by default outputs a more complete and efficient representation of 
the simulated ancestry, but still does not provide the exhaustive
detail required.
Indeed, providing the information required to 
support likelihood calculations originally motivated the addition of
the \texttt{record\_full\_arg}
option to \texttt{msprime}~\citep{baumdicker_efficient_2021}.
Forwards-time simulations that output 
ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} similarly do not
typically contain the level of detail required to compute a likelihood under
the KYF approach.

@@

Although we can associate a likelihood with any ARG in \texttt{tskit} format,
there are some caveats. By treating each edge independently,
our algorithm will still produce a value for ARGs with
the various degrees of completeness to which inheritance
is documented by the different ARG inference methods.
% : e.g., polytomies, absence
% of locally unary coalescence nodes, unpreserved identify of internal nodes
% across local trees, etc.
% However, many of these imprecisions violate the implicit
% assumptions of the likelihood computation.
For example, we assume the 
presence of locally unary nodes to correctly identify the
full extent of all lineages going back through time 
and therefore recombination events, but this information
may be incomplete (e.g., the full simplifed ARG in Fig.~\ref{fig:lik-surface})
or even incorrect.
% Without these unary nodes, our algorithm will detect a recombination
% event along both coalescing lineages in case of a (sub)tree height changes along
% the ARG as both edges would switch parent nodes. 
Locally unary nodes are not a common aspect of current ARG 
research~\citep{wong_general_2023}, but 
are estimated by \texttt{tsinfer}, and can be imputed to high
degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
Determining the effects of various forms of \emph{incorrect} topology on
these likelihoods requires further research.

@@

Our results, while preliminary, suggest that ``full ARGs'', in which 
each recombination event is precisely pinpointed in time, 
contain a significant degree of redundancy in terms of capturing 
information about the generative process. 
It is possible, therefore, that explicitly estimating recombination 
events is not necessary to sample ARGs under the SMC (as ARGweaver
and SINGER do), potentially
leading to further improvements in the scalability of such methods.
Another exciting prospect for ARG inference is the development of 
``hybrid'' inference methods, where hueristic methods such as 
\texttt{tsinfer} or \texttt{ARG-Needle} are used to infer
the details of the recent past in huge datasets such as 
UK Biobank, and passing over to a rigorous model-based
approach for the ancient past.
This is similar in spirit to the widely-used approach 
of simulating complex dynamics of the recent past using 
detailed but slow models and using faster coalescent simulations
in the more distant past, when the assumptions required 
are more justified~\citep{bhaskar_distortion_2014,
kelleher_efficient_2018,haller_tree_2018,nelson_accounting_2020}.
% Maybe we shouldn't bother saying this, but I can't resist the 
% opportunity to say "hey, you should be using tskit!"
% It feels like some PIs aren't getting this message, so need to have
% it spelled out.
Just as forward and backward simulations are useful in different contexts
and can be combined to great effect,
large-scale hueristic and MCMC sampling inference methods have their 
applications, and there may be synergy in their combination. 
Practical combination of methods will require cooperation at the 
data structure level, and the interoperability provided by the 
\texttt{tskit} library [something good].

@@

All scripts and data used for this manuscript are available at
\url{https://github.com/gertjanbisschop/smc-bit-paper},
and an implementation of the algorithm is available at
\url{https://github.com/gertjanbisschop/runsmc}.
\FloatBarrier
\bibliographystyle{plainnat}
\bibliography{paper.bib}

>>>>

\section*{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now strong interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024,nielsen_inference_2025}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Numerous applications taking advantage of this detailed history
in inferred ARGs are now appearing~\citep{
stern_approximate_2019,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
ignatieva_ongoing_2022,
wang_complex_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
osmond_estimating_2021,
huang_estimating_2024,
grundler_geographic_2024,
korfmann_simultaneous_2024,
deraje_inferring_2024,
speidel2025high}
and it seems likely that many more will
follow~\citep{harris_database_2019,harris_using_2023}.
Although there are many different approaches to ARG
inference~\citep{wong_general_2023}, two broad classes of method
have been the focus of most recent interest.

% TODO Maybe swap the order of these two paragraphs around, so it's 
% more natrual to talk about the advantages of model-driven stuff next.
The first class of inference methods are Bayesian approaches
that sample ARGs under a population genetics model such as the
coalescent with recombination~\citep{hudson_properties_1983},
and its approximatation, the Sequentially Markovian
Coalescent~\citep{mcvean_approximating_2005,marjoram_fast_2006}.
ARGweaver~\citep{rasmussen_genome-wide_2014} has been the most widely used and studied,
and until recently consistently out-performed other methods in terms 
of accuracy in a variety of benchmarks~\citep{brandt2022evaluation},
and has succesfully been applied in several
contexts~\citep[e.g.,][]{de_chimpanzee_2016,shriner_whole_2018,hejase_genomic_2020,
stankowski_genetic_2024}. 
The recently-introduced method SINGER takes a similar approach,
and by using an improved Monte Carlo algorithm
promises to be even more accurate and substantially more 
scalable~\citep{deng_robust_2024}. 
The second class of inference methods that has been of recent interest 
are based on heuristics and approximate models.
Relate~\citep{speidel_method_2019},
tsinfer~\citep{kelleher_inferring_2019},
ARG-Needle~\citep{zhang_biobank-scale_2023} 
and Threads~\citep{gunnarsson_scalable_2024}
work on quite different 
principles, but share some common properties. 
Firstly, they are all 
heuristic and approximation driven, favouring computational
scalability over a basis in a statistical model.
Secondly, they all produce a single
ARG as output, inferring a deterministic point estimate.
Thirdly, they regard the precise timing of nodes as a separate
problem, focusing only on the relative ordering of nodes when 
producing the initial ARG. General purpose ARG dating methods are 
appearing~\citep{wohns_unified_2022,deng2025general} and can also improve
dating performance in ARGs sampled from the SMC~\citep{deng2025general}.
Fourthly (and most importantly for the 
purposes of this paper) the ARGs that hueristic methods produce 
lack some of the topological information present in the output of ARGweaver or SINGER. 
This occurs, roughly speaking, because of a basic difference in their
approach to uncertainty.

There is considerable uncertainty in ARG inference, and often a fundamental
lack of information in the sequence data to distinguish different
possibilities. Consider, for example, a case in which we have three
ancestral lineages with identical sequences. We know that there
must have been two coalescence events, but there is no mutational information to 
distinguish their relative ordering. Sampling methods overcome this problem
by randomly choosing an ordering, and uncertainty communicated by the order
changing in the different ARGs sampled from the posterior.
The other approach (used by methods such as tsinfer) is to communicate
the uncertainty structurally by means of a polytomy: we have no
information about the intermediate coalescent event, and so we omit it.
Similarly, there is often a fundamental lack of information about the 
details of recombination events, and heuristic methods (in different ways)
omit these details. In many ways, the significant progress made in
scalability by these heuristic methods is \emph{because} they don't
attempt to precisely reconstruct recombination events.

As there is no mutational
information

(e.g., the timing of recombination events, as discussed below).

and the heuristic methods deal with this uncertainty by omitting these events,
while ARGweaver and SINGER provide a sample from the posterior.
(So, some aspects of the resulting ARGs should be interpreted slightly differently;
see \citet{wong_general_2023} for more discussion.)

@@

Computing the likelihood of an ARG under a model is a fundamental element of
probabilistic inference. The classical Kuhner-Yamato-Felsenstein
\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
works by assigning a probablility to each recombination and common
ancestor event, and to the inter-event  waiting times. 
The likelihood of an 
ARG is then a product over all these three classes of term.
Although the KYF formulation is a natural and elegant way to describe 
the likelihood of an ARG given a complete description of

, such complete detail is not present 
in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
Simulators such as the classical ms
program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
which does not contain sufficient detail for likelihood computation.
The msprime
simulator~\citep{kelleher_efficient_2016,baumdicker_efficient_2021} 
by default outputs a more complete and efficient representation of 
the simulated ancestry, but still does not provide the exhaustive
detail required.
Indeed, providing the information required to 
support likelihood calculations originally motivated the addition of
the \texttt{record\_full\_arg}
option to msprime~\citep{baumdicker_efficient_2021}.
Forwards-time simulations that output 
ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} similarly do not
typically contain the level of detail required to compute a likelihood under
the KYF approach.

@@

Although we can associate a likelihood with any ARG in \texttt{tskit} format,
there are some caveats. By treating each edge independently,
our algorithm will still produce a value for ARGs with
the various degrees of completeness to which inheritance
is documented by the different ARG inference methods.
% : e.g., polytomies, absence
% of locally unary coalescence nodes, unpreserved identify of internal nodes
% across local trees, etc.
% However, many of these imprecisions violate the implicit
% assumptions of the likelihood computation.
For example, we assume the 
presence of locally unary nodes to correctly identify the
full extent of all lineages going back through time 
and therefore recombination events, but this information
may be incomplete (e.g., the full simplifed ARG in Fig.~\ref{fig:lik-surface})
or even incorrect.
% Without these unary nodes, our algorithm will detect a recombination
% event along both coalescing lineages in case of a (sub)tree height changes along
% the ARG as both edges would switch parent nodes. 
Locally unary nodes are not a common aspect of current ARG 
research~\citep{wong_general_2023}, but 
are estimated by tsinfer, and can be imputed to high
degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
Determining the effects of various forms of \emph{incorrect} topology on
these likelihoods requires further research.

@@

Our results, while preliminary, suggest that ``full ARGs'', in which 
each recombination event is precisely pinpointed in time, 
contain a significant degree of redundancy in terms of capturing 
information about the generative process. 
It is possible, therefore, that explicitly estimating recombination 
events is not necessary to sample ARGs under the SMC (as ARGweaver
and SINGER do), potentially
leading to further improvements in the scalability of such methods.
Another exciting prospect for ARG inference is the development of 
``hybrid'' inference methods, where hueristic methods such as 
tsinfer or ARG-Needle are used to infer
the details of the recent past in huge datasets such as 
UK Biobank, and passing over to a rigorous model-based
approach for the ancient past.
This is similar in spirit to the widely-used approach 
of simulating complex dynamics of the recent past using 
detailed but slow models and using faster coalescent simulations
in the more distant past, when the assumptions required 
are more justified~\citep{bhaskar_distortion_2014,
kelleher_efficient_2018,haller_tree_2018,nelson_accounting_2020}.
% Maybe we shouldn't bother saying this, but I can't resist the 
% opportunity to say "hey, you should be using tskit!"
% It feels like some PIs aren't getting this message, so need to have
% it spelled out.
Just as forward and backward simulations are useful in different contexts
and can be combined to great effect,
large-scale hueristic and MCMC sampling inference methods have their 
applications, and there may be synergy in their combination. 
Practical combination of methods will require cooperation at the 
data structure level, and the interoperability provided by the 
\texttt{tskit} library [something good].

@@

All scripts and data used for this manuscript are available at
\url{https://github.com/gertjanbisschop/smc-bit-paper},
and an implementation of the algorithm is available at
\url{https://github.com/gertjanbisschop/runsmc}.
\FloatBarrier
\bibliographystyle{abbrvnat}
\bibliography{paper.bib}
